\bf The forestecology R package for fitting and assessing neighborhood models of the effect of interspecific competition on the growth of trees

### 2021-03-10 paper/forestecology.tex

\bigskip
\begin{abstract}
1. Neighborhood competition models are powerful tools to measure the
effect of interspecific competition. Statistical methods to ease the
application of these models are currently lacking.\\
2. We present the \texttt{forestecology} package providing methods to i)
specify neighborhood competition models, ii) evalulate the effect of
competitor species identity using permutation tests, and iii) measure
model performance using spatial cross-validation. Following
\citet{allen_permutation_2020}, we implement a Bayesian linear
regression neighborhood competition model.\\
3. We demonstrate the package's functionality using data from the
Smithsonian Conservation Biology Institute's large forest dynamics plot,
part of the ForestGEO global network of reseach sites. Given ForestGEO's
data collection protocols and data formatting standards, the package was
designed with cross-site compatibility in mind. We highlight the
importance of spatial cross-validation when interpreting model
results.\\
4. The package features i) \texttt{tidyverse}-like structure whereby
verb-named functions can be modularly ``piped'' in sequence, ii)
functions with standardized inputs/outputs of simple features
\texttt{sf} package class, and iii) an S3 object-oriented implementation
of the Bayesian linear regression model. These three facts allow for
clear articulation of all the steps in the sequence of analysis and easy
wrangling and visualization of the geospatial forestry data.
Furthermore, while the package only has Bayesian linear regression
implemented, the package was designed with extensibility to other
methods in mind.
\end{abstract}

@@

Repeat-censused forest plots offer excellent opportunities to test
neighborhood models of the effect of competition on the growth of trees
(\citet{canham_neighborhood_2004}). Neighborhood models of competition
have been used to: test whether the species identity of a competitor
matters (\citet{uriarte_spatially_2004}); measure species-specific
competition coefficients (\citet{das_effect_2012}
\citet{tatsumi_estimating_2016}); test competing models to see what
structures competitive interactions, e.g.~traits or phylogeny
(\citet{allen_permutation_2020}; \citet{uriarte_trait_2010}); and inform
selective logging practices (\citet{canham_neighborhood_2006}). Although
these are well-described methods, few methods are currently available
for easy application.

We address this shortcoming with the \texttt{forestecology} R package
providing methods and data for forest ecology model fitting and
assessment, available on CRAN
(\url{https://cran.r-project.org/web/packages/forestecology/index.html})
and on GitHub (\url{https://github.com/rudeboybert/forestecology}). The
package is written to model stem radial growth between two censuses
based on neighborhood competition, largely following the methods in
\citet{allen_permutation_2020}.

Let \(i = 1, \ldots, n_j\) index all \(n_j\) trees of ``focal'' species
group \(j\); let \(j = 1, \ldots, J\) index all \(J\) focal species
groups; and let \(k = 1, \ldots, K\) index all \(K\) ``competitor''
species groups. The average annual growth in diameter at breast height
(DBH) \(y_{ij}\) (in centimeters/year) of the \(i^{th}\) tree of focal
species group \(j\) is modeled as

@@

where \(\beta_{0,j}\) is the diameter-independent growth rate for group
\(j\); \(\text{dbh}_{ij}\) is the DBH of the focal tree at the earlier
census for group \(j\) and \(\beta_{\text{dbh},j}\) quantifies its
corresponding growth rate change; \(\text{BA}_{ijk}\) is the sum of the
basal area of all trees of competitor species group \(k\) and
\(\lambda_{jk}\) quantifies the corresponding change in growth for
individuals of group \(j\) from these competitors; and \(\epsilon_{ij}\)
is a random error term distributed \(\text{Normal}(0, \sigma^2)\). They
estimate all parameters via Bayesian linear regression, while exploiting
Normal/Inverse Gamma conjugacy to derive closed-form solutions to all
posterior distributions\footnote{See S1 Appendix of
  \citet{allen_permutation_2020}, available at
  \url{https://doi.org/10.1371/journal.pone.0229930.s004}}. These
closed-form solutions are not as computationally expensive as
approximations from Markov Chain Monte Carlo algorithms.

@@

We present a case-study of \texttt{forestecology}'s functionality on
data from the Smithsonian Conservation Biology Institute (SCBI) large
forest dynamics plot in Front Royal, VA, USA, part of the ForestGEO
global network of research sites
\citep[\citet{andersonteixeira_ctfs-forestgeo_2015}]{bourg_initial_2013}
\citep{bourg_initial_2013}. The 25.6 ha (640 x 400 m) plot is located at
the intersection of three of the major physiographic provinces of the
eastern US---the Blue Ridge, Ridge and Valley, and Piedmont
provinces---and is adjacent to the northern end of Shenandoah National
Park.

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{ggplot}\NormalTok{() }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ growth_scbi }\OperatorTok{%>%}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\DecValTok{500}\NormalTok{), }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{size =}\NormalTok{ growth)) }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{scale_size_binned}\NormalTok{(}\DataTypeTok{limits =} \KeywordTok{c}\NormalTok{(}\FloatTok{0.1}\NormalTok{, }\DecValTok{1}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

We observe 4 variables describing 20 competitor trees: the unique
tree-stem ID, the distance to the focal tree (all \(\leq\) 7.5), the
species, and the basal area (in m\(^2\)) calculated as
\(\frac{\pi \times (\text{DBH/2})^2}{10000}\) for the DBH in cm from the
earlier census. Saving competitor information in list-column format
minimizes redundancy since we do not need to repeat information on the
focal tree 20 times. We visualize the spatial distribution of these
trees in Figure \ref{fig:scbi-focal-vs-comp-map}.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{comp_bayes_lm_scbi}
\CommentTok{## Bayesian linear regression model parameters with a multivariate Normal}
\CommentTok{## likelihood. See ?comp_bayes_lm for details:}
\CommentTok{## }
\CommentTok{##   parameter_type           prior posterior}
\CommentTok{## 1 Inverse-Gamma on sigma^2 a_0   a_star   }
\CommentTok{## 2 Inverse-Gamma on sigma^2 b_0   b_star   }
\CommentTok{## 3 Multivariate t on beta   mu_0  mu_star  }
\CommentTok{## 4 Multivariate t on beta   V_0   V_star   }
\CommentTok{## }
\CommentTok{## Model formula:}
\CommentTok{## growth ~ sp + dbh + dbh * sp + acne * sp + acru * sp + amar * sp + astr}
\CommentTok{## * sp + caca * sp + caco * sp + cade * sp + cagl * sp + caovl * sp + cato}
\CommentTok{## * sp + ceca * sp + ceoc * sp + chvi * sp + cofl * sp + crpr * sp + crsp}
\CommentTok{## * sp + divi * sp + elum * sp + fagr * sp + fram * sp + frni * sp + frpe}
\CommentTok{## * sp + havi * sp + ilve * sp + juci * sp + juni * sp + libe * sp + litu}
\CommentTok{## * sp + nysy * sp + pist * sp + pivi * sp + ploc * sp + prav * sp + prse}
\CommentTok{## * sp + qual * sp + quco * sp + qufa * sp + qumi * sp + qupr * sp + quru}
\CommentTok{## * sp + quve * sp + rops * sp + saal * sp + saca * sp + tiam * sp + ulam}
\CommentTok{## * sp + ulru * sp + unk * sp + vipr * sp}
\end{Highlighting}
\end{Shaded}

@@

For many users the visualizations of \(\lambda_{j,k}\) will be of
particular interest as they provide insight into species competitive
interactions, where negative values indicate a competitor species which
slows the growth of a focal species. Here, for example, we see that
tulip poplars (litu) have a strong negative effect on the growth of
conspecifics but relatively lesser effect on pignut hickory (cagl) and
red oak (quru) neighbors.

@@

\texttt{run\_cv()} also accepts the \texttt{run\_shuffle} argument in
order to permute competitor species identity as described in Section
\ref{permutation-test}. Figure \ref{fig:scbi-simulation} compares model
performance for 49 permutations of competitor species and RMSE
calculations, both with and without cross-validation. Without
cross-validation, competitor species identity does matter as the
observed RMSE was significantly lower than the permutation null
distribution of RMSE. However, once we incorporate spatial
cross-validation, this improvement disappears. These results suggest
that in this 9 ha subplot of the SCBI plot, competitive interactions do
not depend on the identity of the competitor, which is the opposite of
what has been observed in other locations \citep[
\citet{uriarte_spatially_2004}]{allen_permutation_2020}. Furthermore,
the larger cross-validated RMSE's are indicative of the importance of
being vigilant against potential model overfitting.

@@

The \texttt{forestecology} package provides an accessible way to fit and
test models of neighborhood competition. While the package is designed
with ForestGEO plot data in mind, we envision that it can be modified to
work on any single large, mapped forest plot in which at least two
measurements of each individual have been taken, e.g.~the US Forest
Service Forest Inventory and Analysis plots. Furthemore, we also
envision the package being used more generally to model interactions of
any community of mapped sessile organisms \citep{smith_forest_2002}. In
future versions of \texttt{forestecology} we also hope to include models
that account for tree mortality in addition to tree growth. The package
follows the \texttt{tidy} data design principles, leverages the
\texttt{sf} package for spatial data, and S3 open-oriented model
implementation structure. We hope that the package will increase the use
of neighborhood competition models to better understand what structures
plant competition.

>>>>

\bigskip
\begin{abstract}
1. Neighborhood competition models are powerful tools to measure the
effect of interspecific competition. Statistical methods to ease the
application of these models are currently lacking.\\
2. We present the \texttt{forestecology} package providing methods to i)
specify neighborhood competition models, ii) evaluate the effect of
competitor species identity using permutation tests, and iii) measure
model performance using spatial cross-validation. Following
\citet{allen_permutation_2020}, we implement a Bayesian linear
regression neighborhood competition model.\\
3. We demonstrate the package's functionality using data from the
Smithsonian Conservation Biology Institute's large forest dynamics plot,
part of the ForestGEO global network of research sites. Given
ForestGEO's data collection protocols and data formatting standards, the
package was designed with cross-site compatibility in mind. We highlight
the importance of spatial cross-validation when interpreting model
results.\\
4. The package features i) \texttt{tidyverse}-like structure whereby
verb-named functions can be modularly ``piped'' in sequence, ii)
functions with standardized inputs/outputs of simple features
\texttt{sf} package class, and iii) an S3 object-oriented implementation
of the Bayesian linear regression model. These three facts allow for
clear articulation of all the steps in the sequence of analysis and easy
wrangling and visualization of the geospatial data. Furthermore, while
the package only has Bayesian linear regression implemented, the package
was designed with extensibility to other methods in mind.
\end{abstract}

@@

Repeat-censused forest plots offer excellent opportunities to test
neighborhood models of the effect of competition on the growth of trees
\citep{canham_neighborhood_2004}. Neighborhood models of competition
have been used to: test whether the species identity of a competitor
matters {[}\citet{uriarte_spatially_2004}; measure species-specific
competition coefficients \citep[
\citet{tatsumi_estimating_2016}]{das_effect_2012}; test competing models
to see what structures competitive interactions, e.g.~traits or
phylogeny \citep{allen_permutation_2020, uriarte_trait_2010}; and inform
selective logging practices \citep{canham_neighborhood_2006}. Although
these are well-described methods, few methods are currently available
for easy application.

We address this shortcoming with the \texttt{forestecology} R package
providing methods and data for forest ecology model fitting and
assessment, available on CRAN
(\url{https://cran.r-project.org/package=forestecology}) and on GitHub
(\url{https://github.com/rudeboybert/forestecology}). The package is
written to model stem diameter growth between two censuses based on
neighborhood competition, largely following the methods in
\citet{allen_permutation_2020}.

Let \(i = 1, \ldots, n_j\) index all \(n_j\) trees of ``focal'' species
\(j\); let \(j = 1, \ldots, J\) index all \(J\) focal species; and let
\(k = 1, \ldots, K\) index all \(K\) ``competitor'' species. The average
annual growth in diameter at breast height (DBH) \(y_{ij}\) (in
centimeters/year) of the \(i^{th}\) tree of focal species \(j\) is
modeled as

@@

where \(\beta_{0,j}\) is the diameter-independent growth rate of species
\(j\); \(\text{dbh}_{ij}\) is the DBH of the focal tree at the earlier
census and \(\beta_{\text{dbh},j}\) the slope of that species's
diameter-growth relationship; \(\text{BA}_{ijk}\) is the sum of the
basal area of all trees of competitor species \(k\) and \(\lambda_{jk}\)
quantifies the corresponding change in growth for individuals of group
\(j\) from these competitors; and \(\epsilon_{ij}\) is a random error
term distributed \(\text{Normal}(0, \sigma^2)\).
\citet{allen_permutation_2020} estimate all parameters via Bayesian
linear regression, while exploiting Normal/Inverse Gamma conjugacy to
derive closed-form solutions to all posterior distributions\footnote{See
  S1 Appendix of \citet{allen_permutation_2020}, available at
  \url{https://doi.org/10.1371/journal.pone.0229930.s004}}. These
closed-form solutions are not as computationally expensive as
approximations from Markov Chain Monte Carlo algorithms.

@@

We present a case-study of \texttt{forestecology}'s functionality on
data from the Smithsonian Conservation Biology Institute (SCBI) large
forest dynamics plot in Front Royal, VA, USA, part of the ForestGEO
global network of research sites
\citep[\citet{andersonteixeira_ctfs-forestgeo_2015}]{bourg_initial_2013}.
The 25.6 ha (640 x 400 m) plot is located at the intersection of three
of the major physiographic provinces of the eastern US---the Blue Ridge,
Ridge and Valley, and Piedmont provinces---and is adjacent to the
northern end of Shenandoah National Park.

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{ggplot}\NormalTok{() }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ growth_scbi }\OperatorTok{%>%}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\DecValTok{500}\NormalTok{), }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{size =}\NormalTok{ growth)) }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{scale_size_binned}\NormalTok{(}\DataTypeTok{limits =} \KeywordTok{c}\NormalTok{(}\FloatTok{0.1}\NormalTok{, }\DecValTok{1}\NormalTok{)) }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{size =} \KeywordTok{expression}\NormalTok{(}\KeywordTok{paste}\NormalTok{(Growth, }\StringTok{" (cm "}\NormalTok{,y}\OperatorTok{^}\NormalTok{\{}\OperatorTok{-}\DecValTok{1}\NormalTok{\},}\StringTok{')'}\NormalTok{)) )}
\end{Highlighting}
\end{Shaded}

@@

We observe 4 variables describing 20 competitor trees: the unique
tree-stem ID, the distance to the focal tree (all \(\leq\) 7.5 m), the
species, and the basal area (in m\(^2\)) calculated as
\(\frac{\pi \times (\text{DBH/2})^2}{10000}\) for the DBH in cm from the
earlier census. Saving competitor information in list-column format
minimizes redundancy since we do not need to repeat information on the
focal tree 20 times. We visualize the spatial distribution of these
trees in Figure \ref{fig:scbi-focal-vs-comp-map}.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{comp_bayes_lm_scbi}
\CommentTok{## Bayesian linear regression model parameters with a multivariate Normal likelihood. See ?comp_bayes_lm for details:}
\CommentTok{## }
\CommentTok{##   parameter_type           prior posterior}
\CommentTok{## 1 Inverse-Gamma on sigma^2 a_0   a_star   }
\CommentTok{## 2 Inverse-Gamma on sigma^2 b_0   b_star   }
\CommentTok{## 3 Multivariate t on beta   mu_0  mu_star  }
\CommentTok{## 4 Multivariate t on beta   V_0   V_star   }
\CommentTok{## }
\CommentTok{## Model formula:}
\CommentTok{## growth ~ sp + dbh + dbh * sp + acne * sp + acru * sp + amar * sp + astr * sp + caca * sp + caco * sp + cade * sp + cagl * sp + caovl * sp + cato * sp + ceca * sp + ceoc * sp + chvi * sp + cofl * sp + crpr * sp + crsp * sp + divi * sp + elum * sp + fagr * sp + fram * sp + frni * sp + frpe * sp + havi * sp + ilve * sp + juci * sp + juni * sp + libe * sp + litu * sp + nysy * sp + pist * sp + pivi * sp + ploc * sp + prav * sp + prse * sp + qual * sp + quco * sp + qufa * sp + qumi * sp + qupr * sp + quru * sp + quve * sp + rops * sp + saal * sp + saca * sp + tiam * sp + ulam * sp + ulru * sp + unk * sp + vipr * sp}
\end{Highlighting}
\end{Shaded}

@@

For many users the visualizations of \(\lambda_{j,k}\) will be of
particular interest as they provide insight into species-specific
competitive interactions, where negative values indicate a competitor
species which slows the growth of a focal species. Here, for example, we
see that tulip poplars (litu) have a strong negative effect on the
growth of conspecifics but relatively lesser effect on pignut hickory
(cagl) and red oak (quru) neighbors.

@@

\texttt{run\_cv()} also accepts the \texttt{run\_shuffle} argument in
order to permute competitor species identity as described in Section
\ref{permutation-test}. Figure \ref{fig:scbi-simulation} compares model
performance for 49 permutations of competitor species and RMSE
calculations, both with and without cross-validation. Without
cross-validation, competitor species identity does matter as the
observed RMSE was significantly lower than the permutation null
distribution of RMSE. However, once we incorporate spatial
cross-validation, this improvement disappears. These results suggest
that in this 9 ha subplot of the SCBI plot, competitive interactions do
not depend on the identity of the competitor, which is the opposite of
what has been observed in other locations \citep[
\citet{uriarte_spatially_2004}]{allen_permutation_2020}. This provides a
striking example of the importance of cross-validation, as without it
the over-fit model gives rise to an incorrect conclusion.

@@

The \texttt{forestecology} package provides an accessible way to fit and
test models of neighborhood competition. While the package is designed
with ForestGEO plot data in mind, we envision that it can be modified to
work on any single large, mapped forest plot in which at least two
measurements of each individual have been taken. Furthermore, we also
envision the package being used more generally for inventory
plot-structure with many spatially separated plots, e.g.~the US Forest
Service Forest Inventory and Analysis plots \citep{smith_forest_2002}.
In future versions of \texttt{forestecology} we also hope to include
models that account for tree mortality in addition to tree growth. The
package follows the \texttt{tidy} data design principles, leverages the
\texttt{sf} package for spatial data, and S3 open-oriented model
implementation structure. We hope that the package will increase the use
of neighborhood competition models to better understand what structures
plant competition.
